GPU-based Adaptive Beamforming for Active Sonar Imaging

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	23:20:50	(Mon 7.May)

Consider an $M$ element phased array. Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights be  $w_m[n]$ and $\Delta_m[n]$. To simplify the upcoming notation, let sample $x_m[n]$ delayed by a factor $\Delta_m$ be written as:
\begin{align}
x^\Delta_m = x_m[n-\Delta_m[n]].
\end{align}
The beamformer output $z[n]$ is then simply the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x^\Delta_0[n]\\x^\Delta_1[n]\\\vdots\\x^\Delta_{M-1}[n]}.\label{z}
\end{align}
The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction. The solution to this optimization problem is given as:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x[n]\H\}$ is the spatial covariance matrix. Since $\R$ is unknown, it must be estimated. We will do this by computing the sample covariance matrix, $\eR$. In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to achieve \gls{DAS}-like speckle statistics, and \emph{diagonal loading} to ensure a numerically well conditioned $\eR$~\cite{Synnevag2009a}.

@@

In the upcoming results, we have beamformed pre-delayed data. This is a large dataset, and the latency experienced when copying it from the \gls{CPU} side to the \gls{GPU} was significant. However, in a practical system only the channel data should be transferred as the \gls{GPU} can effortlessly also perform simple delaying of data. The remaining latency can be easily hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.

@@

TODO:
\begin{itemize}
\item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing \gls{MVDR} 
\item Introduce CUDA, or ditch it?
\item Doesn't sell well that we used nVidias solver, especially when that's the bottleneck in the design...
\end{itemize}

@@

We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from the 32 element Kongsberg Maritime HISAS1030 synthetic aperture sonar~\cite{Hansen2009}. The sonar was attached to the HUGIN \gls{AUV} and used in strip-map mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}). Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth. The \gls{MVDR} image was created with the parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{gfx/benchmark.pdf}
\caption{\gls{MVDR} benchmarks. \protect\newline A 55\,kpx image processed with $M=32$ channels and $d=2\%$ diagonal loading.}\label{benchmarks}
\end{figure}

To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (\Fig{benchmarks}). For this test we used a 55\,kpixels sectorscan image. The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card. This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops. We kept the channel count at $M=32$ and added $d=2\%$ diagonal loading.

>>>>

Consider an $M$ element phased array. Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights be  $w_m[n]$ and $\Delta_m[n]$. To simplify the upcoming notation, let sample $x_m[n]$ delayed by a factor $\Delta_m$ be written as:
\begin{align}
x^\Delta_m = x_m[n-\Delta_m[n]].
\end{align}
The beamformer output $z[n]$ is then simply the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x^\Delta_0[n]\\x^\Delta_1[n]\\\vdots\\x^\Delta_{M-1}[n]}.\label{z}
\end{align}
If the weights in (\ref{z}) were static, this would be referred to as the \gls{DAS} beamformer. The \gls{MVDR} beamformer~\cite{Capon1969}, however, finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction. The solution to this optimization problem is given as:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x[n]\H\}$ is the spatial covariance matrix. Since $\R$ is unknown, it must be estimated. We will do this by computing the sample covariance matrix, $\eR$. In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to achieve \gls{DAS}-like speckle statistics, and \emph{diagonal loading} to ensure a numerically well conditioned $\eR$~\cite{Synnevag2009a}.

@@

In the upcoming results, we have beamformed pre-delayed data. This is a large dataset, and the latency experienced when copying it from the \gls{CPU} side to the \gls{GPU} was significant. However, in a practical system only the channel data should be transferred as the delaying of data can effortlessly be performed by the \gls{GPU}. The remaining latency can be hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.

In this paper we have targeted an nVidia \gls{GPU} using the C interface to nVidia's \gls{CUDA} framework~\cite{NvidiaCuda}.

@@

@@

We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from the 32 element Kongsberg Maritime HISAS1030~\cite{Hansen2009}. The sonar was attached to the HUGIN \gls{AUV} and used in sidescan mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}). Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth. The \gls{MVDR} image was created with the parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{gfx/benchmark.pdf}
\caption{\gls{MVDR} benchmarks. Forming a 55\,kpixels image from a $M=32$ channel array.}\label{benchmarks}
\end{figure}

To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (\Fig{benchmarks}). For this test we used a 55\,kpixels sectorscan image. The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card. This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
